Models

### 2021-02-01 analysis/05_model_metrics.Rmd
Fit models based on numer of plots within study.

metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  select(-cYEAR) %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
```
####Data exploration

@@

Fit GLM to overall metric trend
```{r}
# model for overall trend (random effect w/varying intercept)
## should be GLS according to Dornelas supplement, but glmer isn't happy
metric_mean_fit <- metrics %>%
  filter(!is.na(value)) %>%
  mutate(minyear = min(year), maxyear = max(year)) %>%
  group_nest(metric, minyear, maxyear) %>%
  #mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), data = .x)),
  mutate(model = map(data, ~lmer(logvalue ~ year + (1|rarefyID), data = .x)),
         prediction = map(model, ~predict(.x), newdata = as.data.frame(year = minyear:maxyear)))

@@

mean_tests <- metric_mean_fit %>%
  #filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
   mutate(p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull())) %>% 
  select(metric, p_value) %>%
  left_join(metric_mean_coef %>% 
              filter(term == "year") %>% 
              select(metric, estimate))

# model for overall trend (random effect w/varying intercept and slope)
# metric_slope_fit <- metrics %>%
#   filter(!is.na(value)) %>%
#   group_nest(metric) %>%
#   mutate(model = map(data, ~lmer(logvalue ~ year + (1 + year | rarefyID), data = .x)))
# 
# metric_slope_coef <- metric_slope_fit %>%
#   mutate(coef = map(model, tidy)) %>% 
#   #mutate(confint = map(model, confint)) %>%
#   select(-data, -model) %>% 
#   unnest(cols = c(coef))
```

```{r}
indiv_ts_fit <- metrics %>% 
  filter(!is.na(value)) %>%
  group_by_at(setdiff(names(metrics), c("year", "value", "logvalue"))) %>%
  filter(n_distinct(year) > 1) %>% #filter out rarefyID's that don't have multiple years after dropping Na's
  group_nest() %>%
  mutate(model = map(data, ~ lm(logvalue ~ year, data = .x)))

indiv_ts_coef <- indiv_ts_fit %>%
  mutate(coef = map(model, tidy)) %>% 
  select(-data, -model) %>% 
  unnest(cols = c(coef))
```

>>>>

metrics <- pins::pin_get("rarefied-medians", board = "github") %>%
  select(-cYEAR) %>%
  pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + abs(floor(min(na.omit(value)))))) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  filter(!is.na(value))
```
####Data exploration

@@

Fit GLM to overall metric trend
```{r}
# model for overall trend (random effect w/varying intercept)
## should be GLS according to Dornelas supplement, but glmer isn't happy
metric_mean_fit <- metrics %>%
  mutate(minyear = min(year), maxyear = max(year)) %>%
  group_nest(metric, minyear, maxyear) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year + (1|study_id) + (1|rarefyID), 
                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))

@@

mean_tests <- metric_mean_fit %>%
  #filter(metric %in% c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
   mutate(p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull())) %>% 
  select(metric, p_value) %>%
  left_join(metric_mean_coef %>% 
              filter(term == "year") %>% 
              select(metric, estimate))
```

Fit model to individual studies - if there is more than one timeseries in a study, account for that
```{r}
sing_study <- metrics %>%
  group_by(study_id) %>%
  filter(n_distinct(rarefyID) == 1) %>% 
  pull(study_id) %>% unique()

# indiv_data <- metrics %>%
#   mutate(rarefyID = as.factor(rarefyID)) %>%
#   group_by_at(setdiff(names(metrics), c("year", "value", "logvalue", "rarefyID", "cell", "samplen"))) %>% 
#   #filter(n_distinct(year) > 1) %>% #filter out rarefyID's that don't have multiple years after dropping Na's
#   group_nest()

indiv_data <- metrics %>%
  mutate(rarefyID = as.factor(rarefyID)) %>%
  group_by(study_id) %>%
  group_nest()

indiv_ts_sing_fit <- indiv_data %>%
  filter(study_id %in% sing_study) %>%
  mutate(model = map(data, ~ lm(logvalue ~ year, data = .x))) %>%
  mutate(model_type = "lm")

indiv_ts_fit <- indiv_data %>%
  filter(!study_id %in% sing_study) %>%
  mutate(model = map(data, ~lmer(logvalue ~ (1 + year|rarefyID), 
                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead")))) %>%
  mutate(model_type = "mixed",
         p_value = map(model, ~drop1(.x, test = "Chisq") %>% select(`Pr(Chi)`) %>% slice(2) %>% pull()))

indiv_ts_coef <- bind_rows(indiv_ts_fit, indiv_ts_sing_fit) %>%
  mutate(coef = map(model, tidy)) %>% 
  select(-data, -model) %>% 
  unnest(cols = c(coef))
```

```{r}
fit_mod <- function(data){
  print(unique(data$rarefyID)) %>%
  lmer(logvalue ~ (1 + year|rarefyID), 
                                 data = data, control = lmerControl(optimizer ="Nelder_Mead"))
}

indiv_data %>%
  filter(!study_id %in% sing_study) %>%
  mutate(model = map(data, ~lmer(logvalue ~ ( 1+ year|rarefyID), 
                                 data = .x, control = lmerControl(optimizer ="Nelder_Mead"))))

fit_mod <- function(filter_id){
  filter_data <- metrics %>% filter(study_id == filter_id)

  print(filter_id)
  lmer(logvalue ~ (1+ year|rarefyID), 
                                 data = filter_data)
}

mult_studies <- metrics %>% filter(!study_id %in% sing_study) %>% pull(study_id) %>% unique()
map(mult_studies, fit_mod)
